Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-05 tagless/tagless-final.tex
small remarks

We may now make |P| an instance of
|Symantics| and implement the partial evaluator as follows. We elide
|mul|, |leq|, |if_|, and |fix|.
\begin{code}
instance Symantics P where
  int x  = VI x
  bool b = VB b
  add (VI n1) (VI n2) = VI (n1 + n2)
  add e1 e2 = E (add (abstr e1) (abstr e2))
  lam = VF
  app (VF f) ea = f ea
  app (E f)  ea = E (app f (abstr ea))
\end{code}
The implementations of |int|, |bool|, and |add| are like
in~\S\ref{S:PE-problem}.
The interpretation of |lam f| just wraps the
HOAS function |f| into the result. We can always compile |f| to a code value,
but we delay it to apply |f| to concrete arguments. The interpretation of
|app ef ea| checks to see if |ef| is such a delayed
HOAS function |VF f|. If it is, we apply |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
dynamic value |E f|, we residualize.

@@

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code}
repr_pe Int      = Int
repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and more portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new |Symantics| signature.

@@

\oleg{He did not eliminate all the tags.}\jacques{can you show me?}
From reading \citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}, our 
impression is that the claim is:
\begin{enumerate}
\item One needs to encode a typed language in a typed language based on
a sum type,
\item It is not possible to write a \emph{direct} interpreter for an typed
language encoded this way, in a typed language, without either using a
very advanced type system or incurring tagging/untagging overhead.
\item Thus we have to use an indirect interpreter, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags.
\item Hence the self-interpreter cannot be Jones-optimal.
\end{enumerate}
While the logic is sound, we showed that the premise in the first step
was not valid.

\oleg{the only known to them, and \emph{partial} solution, was 
  that by Zhe Yang;}\jacques{I cannot find that reference}

@@

Typing Dep Typing plus two messages on Haskell-Cafe: typechecking 
from untyped to the typed form, ready to be interpreted. Walid's ICFP02
uses dependent types

@@

Our approach showed better the meaning of GADT and when they can be
avoided.
This shows also how a limited form of GADTs (the ones
used for the sake of type-dependent representations) can be
implemented with just functions.

separate compilation: separate
typechecking of terms without knowing which particular
hole they will be plugged into. Formulate this as a prop?

>>>>

We may now make |P| an instance of
|Symantics| and implement the partial evaluator as follows. We elide
|mul|, |leq|, |if_|, and |fix|.
\begin{code}
instance Symantics P where
  int x  = VI x
  bool b = VB b
  add (VI n1) (VI n2) = VI (n1 + n2)
  add e1 e2 = E (add (abstr e1) (abstr e2))
  lam = VF
  app (VF f) ea = f ea
  app (E f)  ea = E (app f (abstr ea))
\end{code}
The implementations of |int|, |bool|, and |add| are like
in~\S\ref{S:PE-problem}.
The interpretation of |lam f| just wraps the
HOAS function |f|. We can always compile |f| to a code value,
but we delay it to apply |f| to concrete arguments. The interpretation of
|app ef ea| checks to see if |ef| is such a delayed
HOAS function |VF f|. If it is, we apply |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example |testpowfix7| in~\S\ref{S:PE-solution}). If |ef| is a
dynamic value |E f|, we residualize.

@@

\subsection{The ``final'' solution}
\label{S:PE-solution}
Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
data P t = P (Maybe (repr_pe t)) (C t)
\end{code}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code}
repr_pe Int      = Int
repr_pe Bool     = Bool
repr_pe (a -> b) = P a -> P b
\end{code}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. Switching to MetaOCaml to demonstrate this
portability, we recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new |Symantics| signature.

@@

\oleg{He did not eliminate all the tags.}\jacques{can you show me?}
\oleg{WalidICFP02, the very end of p1 and the beginning of p2, first
  column}
From reading \citet{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple}, our 
impression is that the claim is:
\begin{enumerate}
\item One needs to encode a typed language in a typed language based on
a sum type,
\item It is not possible to write a \emph{direct} interpreter for an typed
language encoded this way, in a typed language, without either using a
very advanced type system or incurring tagging/untagging overhead.
\item Thus we have to use an indirect interpreter, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags.
\item Hence the self-interpreter cannot be Jones-optimal.
\end{enumerate}
While the logic is sound, we showed that the premise in the first step
was not valid.

From the above:
For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''

@@

Typing Dep Typing plus two messages on Haskell-Cafe: typechecking 
from untyped to the typed form, ready to be interpreted. Walid's ICFP02
uses dependent types
\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019160.html}
\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019161.html}
\url{http://www.haskell.org/pipermail/haskell-cafe/2006-October/019168.html}

@@

separate compilation: separate
typechecking of terms without knowing which particular
hole they will be plugged into.
